---
title: "figure6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/henry/TWAS_for_protein")
```

Code for generating figure 6 in "Transcriptome-wide association study of the plasma proteome reveals cis and trans regulatory mechanisms underlying complex traits."

```{r}
library(data.table)
library(dplyr)
library(tidyr)
library(qvalue)
library(ggplot2)
library(viridis)
library(lattice)
library(grid)
library(gridExtra)
library(ggpubr)
library(eulerr)
library(ggrepel)
library(forcats)
"%&%" = function(a,b) paste(a,b,sep="")
```

# Figure 6a Number of Significant Associations Observed vs Predicted Expression
```{r}
all_sig_cis_same_results <- fread("output/3-9-2023/all_signficant_FDR0.05_cis_same_results_TOPMed.txt")

all_unique_sig_predicted_results <- all_sig_cis_same_results %>% 
  filter(expression_type == "predicted") %>% 
  distinct(transcript_ENSG, .keep_all = T)
unique_predicted_genes <- all_unique_sig_predicted_results$transcript_ENSG

all_unique_sig_observed_results <- all_sig_cis_same_results %>% 
  filter(expression_type == "observed") %>% 
  distinct(transcript_ENSG, .keep_all = T)
unique_observed_genes <- all_unique_sig_observed_results$transcript_ENSG

intersecting_genes <- intersect(unique_predicted_genes, unique_observed_genes)

#sig_associations_counts <- data.frame(expression_type = c("observed","predicted", "observed", "predicted"), num_associations = c(length(unique_observed_genes) - length(intersecting_genes), length(unique_predicted_genes) - length(intersecting_genes), length(intersecting_genes), length(intersecting_genes)), unique = c("unique","unique","shared","shared"))
  
sig_associations_counts <- data.frame(expression_type = c("Observed","Predicted"), num_associations = c(length(unique_observed_genes) - length(intersecting_genes), length(unique_predicted_genes) - length(intersecting_genes)))

venndiag <- euler(c(Obs = length(unique_observed_genes) - length(intersecting_genes),
                    Pred = length(unique_predicted_genes) - length(intersecting_genes),
                    "Pred&Obs" = length(intersecting_genes)))

fig6a <- ggplot(sig_associations_counts, aes(x=expression_type, y=num_associations, fill=expression_type)) +
  geom_bar(stat="identity") +
  scale_fill_manual(values = c("#de4968","#414487"), name="Expression Type", labels = c("Observed","Predicted")) +
    theme_classic(10) + 
    theme(axis.text = element_text(size = 10), 
          axis.text.x = element_text(),
          axis.title = element_text(size = 11)) +
    labs(x=element_blank(),y="Unique Associations")
fig6a
```

# Figure 6b Pi1s Observed vs Predicted Expression
```{r}
predicted_pi1s <- fread("output/3-9-2023/TOPMed_pred_expr_all_cis_same_pi1s_lambda0.75.txt")
observed_pi1s <- fread("output/3-9-2023/TOPMed_obs_expr_all_cis_same_pi1s_lambda0.75.txt")

cis_trans <- c(rep("trans_acting",49),rep("cis_acting",49),rep("cis_same",49),rep("cis_different",49))
groups <- as.data.frame(cis_trans)
pred_obs <- c(rep("predicted",49*4))
expression <- as.data.frame(pred_obs)
tissue <- c(rep(predicted_pi1s$tissue, 4))
tissues <- as.data.frame(tissue)
pi1s <- c(predicted_pi1s$trans_acting, predicted_pi1s$cis_acting, predicted_pi1s$cis_same, predicted_pi1s$cis_different)
values <- as.data.frame(pi1s)
pred_pi1s_for_plotting <- cbind(expression,tissues,groups,values)
pred_pi1s_for_plotting <- pred_pi1s_for_plotting %>% filter(cis_trans == "cis_same")

cis_trans <- c(rep("trans_acting",3),rep("cis_acting",3),rep("cis_same",3),rep("cis_different",3))
groups <- as.data.frame(cis_trans)
pred_obs <- c(rep("observed",3*4))
expression <- as.data.frame(pred_obs)
tissue <- c(rep(observed_pi1s$tissue, 4))
tissues <- as.data.frame(tissue)
pi1s <- c(observed_pi1s$trans_acting, observed_pi1s$cis_acting, observed_pi1s$cis_same, observed_pi1s$cis_different)
values <- as.data.frame(pi1s)
obs_pi1s_for_plotting <- cbind(expression,tissues,groups,values)
obs_pi1s_for_plotting <- obs_pi1s_for_plotting %>% filter(cis_trans == "cis_same")

cis_same_pi1s_for_plotting <- rbind(pred_pi1s_for_plotting, obs_pi1s_for_plotting)

fig6b <- ggplot(pred_pi1s_for_plotting, aes(x=pi1s)) + 
  geom_density(alpha = 0.5, fill = "#414487") +
  scale_x_continuous(limits = c(0, 0.6)) +
  geom_vline(xintercept=max(obs_pi1s_for_plotting$pi1s),size=1, color="#de4968") +
  annotate("text", x=(max(obs_pi1s_for_plotting$pi1s)-0.13), y=4, label="PBMC: "%&%round(max(obs_pi1s_for_plotting$pi1s),3), size=4, color="#de4968") +
  #scale_x_continuous(limits = c(0, 1.0)) +
  theme_classic(10) + 
  theme(axis.text = element_text(size = 10), 
        axis.text.x = element_text(angle = 45, hjust=1),
        axis.title = element_text(size = 11),
        legend.title = element_text(size=10),
        legend.text = element_text(size=9)) +
  labs(x=expression(paste("Distribution of Cis-same ",pi[1],"'s across Tissues")),y=element_blank())
fig6b
```

# Figure 6c Comparing Correlation of Expression with Protein Abundance
```{r}
max_corr_vals <- fread("output/4-24-2023/significant_cis_same_genes_expression_protein_levels_max_correlations_matrix.txt") 
filtered_max_corr_vals <- max_corr_vals %>% filter(observed_expression != -Inf) %>%
  filter(predicted_expression != -Inf)

fig6c <- ggplot(filtered_max_corr_vals, aes(x=observed_expression, y=predicted_expression)) +
  geom_point() +
  geom_abline(color = "#de4968") +
  theme_classic(10) +
  labs(x="Correlation of Observed Expression\n with Protein Levels",
       y="Correlation of Predicted Expression\n with Protein Levels") +
  theme(axis.text = element_text(size = 10), 
        axis.title = element_text(size = 11))
fig6c
```

# Figure 6d Comparing Correlation of Expression with Protein Abundance
```{r}
#Violin plot?
#Need to reshape so that we have 1 column: predicted or observed expression
#second column is vector of correlation values
obs_correlations <- as.data.frame(filtered_max_corr_vals$observed_expression) %>% mutate(obs_pred = "Observed")
colnames(obs_correlations) <- c("corr","obs_pred")

pred_correlations <- as.data.frame(filtered_max_corr_vals$predicted_expression) %>% mutate(obs_pred = "Predicted")
colnames(pred_correlations) <- c("corr","obs_pred")

corr_vals_for_plotting <- rbind(obs_correlations, pred_correlations)
fig6d <- ggplot(corr_vals_for_plotting, aes(x=obs_pred, y=corr)) +
  geom_violin(position = position_dodge(width = 0.4), width=1, aes(fill=obs_pred), show.legend=F) +
  scale_fill_manual(values = c("#de4968","#414487")) +
  geom_boxplot(width=0.1, position = position_dodge(width = 0.4)) + 
  theme_classic(10) +
  theme(axis.text = element_text(size = 10), 
          axis.text.x = element_text(),
          axis.title = element_text(size = 11)) +
  labs(x=element_blank(),y="Correlation of Expression with Protein Levels")
fig6d
```


