---
title: "building_the_pipeline"
author: "hwittich"
date: "2021-02-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
Now that I have successfully run the PrediXcan software and found significant associations between SNPs on chromsome 22 of individuals in the AFA population and their levels of the protein, SLO11772_ENSG00000100029.18, the next step is to perform the same analysis on a bigger set of data, building a pipeline in the process. All of this will be handled in the python wrapper:   
-- protein_association.py
```{}
python3 protein_association.py --dosages /home/hwittich/mount/TWAS_for_protein/TOPMed_Test_Data/nohead_hg38chr22.maf0.01.R20.8.dosage.txt --exp /home/hwittich/mount/TWAS_for_protein/TOPMed_Test_Data/en_Whole_Blood.db --proteins /home/hwittich/mount/TWAS_for_protein/TOPMed_Test_Data/test_prot_expression.txt --samples /home/hwittich/mount/TWAS_for_protein/TOPMed_Test_Data/samples.txt 
```

# Analyzing results
So, I've run PrediXcan on every protein using the Whole_Blood GTeX model in the ALL population. First, all of our p-values are divided across 1,279 files, separated by protein. Let's combine them to analyze all of our associations at once with the script:
-- concatenate_results.py   

Now let's look at our results:
```{r}
protein_associations <- read.table("/home/hwittich/mount/TWAS_for_protein/output/3-22-2021/Whole_Blood_every_protein_association.txt",header = TRUE, sep="\t")
library(qqman)
hist(protein_associations$pvalue)
qq(protein_associations$pvalue)
```
Great, looking at our histogram, it is clear that we have some significant hits. THis is further supported by our qqplot, which shows that most of the SNPs follow the identitity line, while a handful of SNPs are significant.

Let's subset the significant SNPs.
```{r}
library(dplyr)
library(tidyr)
#First, calculate the bonferroni threshold
#=0.5/((#proteins tested * #transcripts predicted)/total # of tests)
n_proteins<-1279 #length of protein matrix (sort of, duplicates were removed)
n_transcripts<-6689 #length of association output
alpha<-0.05
bonferroni_threshold=alpha/(n_proteins*n_transcripts)
sig_hits <- filter(protein_associations,pvalue<bonferroni_threshold) %>% separate(gene,sep="\\.",into=c("gene_id","gene_version"))
```
Now, for every significant hit, we want to compare the predicted transcript and the protein it is associated with. By comparing the location of each, we can determine if we have a cis-acting or trans-acting hit.
```{r}
#Let's pull in the chromosomal coordinates of the predicted transcript
gene_annotation <- read.table("/home/hwittich/mount/TWAS_for_protein/TOPMed_ALL_data/gene_annotation_hg38_sanitized_BP_CHROM.txt",header=T,sep=" ") %>% select(gene_id,chrchr,start)
sig_hits <- left_join(sig_hits, gene_annotation, by = c("gene_id"="gene_id"), copy=TRUE) %>% rename("gene_chr"="chrchr","gene_start"="start")

#Now let's pull in the chromosomal coordinates of the associated protein
protein_annotation <- read.table("/home/hwittich/mount/TWAS_for_protein/TOPMed_ALL_data/somascan1.3k_gencode.v32_annotation.txt",header=T,sep="\t") %>% select(SomaId,ENSG_id,chr,start) %>% rename("protein_ID"="ENSG_id","protein_chr"="chr","protein_start"="start")
sig_hits <- left_join(sig_hits,protein_annotation, by = c("protein"="SomaId"), copy=TRUE)

#Label the multigene aptamers
multigene_aptamers <- read.table("/home/hwittich/mount/TWAS_for_protein/TOPMed_ALL_data/multigene_aptamers.txt",header = T) %>% separate(joint_ID,sep="_",into=c("SL_ID","ENSG_ID"))
sig_hits <- mutate(sig_hits, multigene = ifelse(protein%in%multigene_aptamers$SL_ID,TRUE,FALSE))
```
Next, we have to correct the positions of each gene so that they're relative to the whole genome, not just the chromosome they're on.
```{r}
chromosomes<-c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10","chr11","chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20","chr21","chr22","chrX","chrY")
lengths<-c(248956422,242193529,198295559,190214555,181538259,170805979,159345973,145138636,138394717,133797422,135086622,133275309,114364328,107043718,101991189,90338345,83257441,80373285,58617616,64444167,46709983,50818468,156040895,57227415)
#Calculate chr start positions by summing length of prior chromosomes
starts <- rep(NA,24)
sum=0
for(i in 1:24){
  starts[i] = sum
  sum = sum + lengths[i]
}
chr_len<-data.frame(chromosomes,starts)

#Now join the lengths to the sig_hits dataframe
sig_hits<-left_join(sig_hits,chr_len,by=c("gene_chr"="chromosomes"),copy=TRUE) %>% rename("gene_chr_start"="starts")
sig_hits<-left_join(sig_hits,chr_len,by=c("protein_chr"="chromosomes"),copy=TRUE) %>% rename("protein_chr_start"="starts")

#Finally, add the chromosome starting position to the gene starting positions
sig_hits<-mutate(sig_hits,new_protein_start = protein_chr_start + protein_start) %>% mutate(new_gene_start = gene_chr_start+gene_start)

#Then, let's compare the positions to determine if a gene is trans-acting or cis-acting
sig_hits<-mutate(sig_hits,cis_trans = ifelse((new_gene_start-1000000)<=new_protein_start & new_protein_start<=(new_gene_start+1000000),"cis-acting","trans-acting"))
```

Now let's plot the chromosomal position of the transcript vs the chromosomal position of the associated protein for every significant hit!
```{r}
library(ggplot2)
ggplot(sig_hits,aes(x=new_gene_start,y=new_protein_start,size=pvalue,color=cis_trans,group=multigene)) + 
  geom_point(aes(shape=multigene)) + 
  scale_shape_manual(values=c(21,22)) +
  scale_size_continuous(guide=FALSE) + 
  scale_color_manual(values=c("dark gray","dark red"),na.value="dark gray") +
  geom_hline(yintercept=starts[1:22],size=0.1) + 
  geom_vline(xintercept=starts[1:22],size=0.1) + 
  coord_cartesian(xlim=c(starts[1],starts[22]),ylim=c(starts[1],starts[22])) + 
  theme_classic(10) + 
  xlab("Gene") + 
  ylab("Target Protein") +
  scale_x_continuous(breaks=starts[1:22],labels=c(1:14,"",16,"",18,"",20,"",22)) + 
  scale_y_continuous(breaks=starts[1:22],labels=c(1:16,"",18,"",20,"",22))
```

